The benefits of reporting critical effect size values

### 2024-08-21 paper/paper.Rmd
m

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The "critical effect size values" refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, which should subsequently be validated by "replication science".

@@

There are two clear benefits to reporting the critical effect size value for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting the critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the 'crud factor' [@orben_crud_2020].

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [@transue2019apa] or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

An advantage of reporting the critical effect size value is that it can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical effect size value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical effect size value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's $d > 0.07$ would reach significance, is statistical significance enough to signal a "successful" treatment? Maybe yes, even if effects are tiny [e.g., @funder2019evaluating], but knowing the critical effect size value certainly prompts some appropriate interpretive caution.

@@

Beyond enhancing study design and statistical inferences based on hypothesis tests, reporting critical effect size value(s) can also serve an educational purpose. It underscores how the distinction between a significant and non-significant result is not solely determined by the presence or absence of a true effect, but also by the sample size. By highlighting a critical effect size value, researchers can become more aware of the possibility of Type 2 errors when results are non-significant. Conversely, in studies with exceedingly large samples and in many meta-analyses, the critical effect size value(s) may serve as a reminder that any observed effect larger than a trivially small value will likely achieve significance. This emphasizes that the mere attainment of statistical significance in a test is not particularly surprising, especially in non-experimental studies.

>>>>

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The "critical effect size value(s)" refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, which should subsequently be validated by "replication science".

@@

There are two clear benefits to reporting the critical effect size value(s) for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the 'crud factor' [@orben_crud_2020].

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [@transue2019apa] or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

An advantage of reporting critical effect size values is that they can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical effect size value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical effect size value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's $d > 0.07$ would reach significance, is statistical significance enough to signal a "successful" treatment? Maybe yes, even if effects are tiny [e.g., @funder2019evaluating], but knowing the critical effect size value certainly prompts some appropriate interpretive caution.

@@

Beyond enhancing study design and statistical inferences based on hypothesis tests, reporting critical effect size values can also serve an educational purpose. It underscores how the distinction between a significant and non-significant result is not solely determined by the presence or absence of a true effect, but also by the sample size. By highlighting a critical effect size value, researchers can become more aware of the possibility of Type 2 errors when results are non-significant. Conversely, in studies with exceedingly large samples and in many meta-analyses, the critical effect size value(s) may serve as a reminder that any observed effect larger than a trivially small value will likely achieve significance. This emphasizes that the mere attainment of statistical significance in a test is not particularly surprising, especially in non-experimental studies.

### 2024-08-21 paper/paper.tex
m

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The ``critical effect size values'' refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample (Lakens, 2022). To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with \(n = 20\) participants and uses \(p < 0.05\) as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's \(r = -0.44\) and \(r = 0.44\). As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size (\(n\)) \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance (Transue, 2019). Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, Wilson et al. (2020) propose that it could serve as a useful preliminary filter utilized by ``original science'' for screening potentially interesting effects, which should subsequently be validated by ``replication science''.

@@

There are two clear benefits to reporting the critical effect size value for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting the critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the `crud factor' (Orben \& Lakens, 2020).

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (Transue, 2019) or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

An advantage of reporting the critical effect size value is that it can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical effect size value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical effect size value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's \(d > 0.07\) would reach significance, is statistical significance enough to signal a ``successful'' treatment? Maybe yes, even if effects are tiny (e.g., Funder \& Ozer, 2019), but knowing the critical effect size value certainly prompts some appropriate interpretive caution.

>>>>

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The ``critical effect size value(s)'' refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample (Lakens, 2022). To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with \(n = 20\) participants and uses \(p < 0.05\) as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's \(r = -0.44\) and \(r = 0.44\). As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance (Transue, 2019). Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, Wilson et al. (2020) propose that it could serve as a useful preliminary filter utilized by ``original science'' for screening potentially interesting effects, which should subsequently be validated by ``replication science''.

@@

There are two clear benefits to reporting the critical effect size value(s) for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the `crud factor' (Orben \& Lakens, 2020).

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (Transue, 2019) or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, critical effect size values can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

An advantage of reporting critical effect size values is that they can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical effect size value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical effect size value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's \(d > 0.07\) would reach significance, is statistical significance enough to signal a ``successful'' treatment? Maybe yes, even if effects are tiny (e.g., Funder \& Ozer, 2019), but knowing the critical effect size value certainly prompts some appropriate interpretive caution.
